\chapter*{Take the Next Step}\label{chap-take-the-next-step}

The undeniable message of the \textit{Bhagavad Gītā} is that our true being, and our eternal relationship to the Supreme Infinite, is obscured by our identity with the body and mind.

Kṛṣṇa tells us our goal is communion with God and the way is \textit{bhakti}, pure devotion, which shifts us from limiting selfishness to unlimited Love. But how do we take this truth to heart and have the courage to walk into the unknown? Only by opening the heart and purifying the mind. 

To help you accomplish this, Paramahamsa Vishwananda offers a holistic path of devotion that is rooted in time-honored spiritual traditions. This \textit{bhakti} path naturally helps you cultivate a uniquely personal, loving relationship with the Lord Himself.

Through His international Bhakti Marga organization, you are invited to take part in a wide variety of programs and ways to learn, all of which will help you to rapidly advance on your spiritual path. Using a unique four-armed approach, Bhakti Marga helps everyone rise in love with God by strengthening faith and increasing devotion. In this way, you can grow and progress in a way that best suits you.

Bhakti Marga's four arms include:
\begin{itemize}[itemsep=1pt, topsep=6pt]

\item Philosophical Knowledge that directs the
mind toward higher truths.

\item Yoga \& Meditation that offer a variety of
meditative practices to awaken the wisdom
of the soul.

\item Rituals that ignite devotion, reconnecting the
soul with the Divine.

\item Devotional Arts that channel the natural
tendency towards pleasure and expression
to attain the highest happiness and loving
communion with God.
\end{itemize}

Sincerely engaging in these practices over time allows you to experience the transformational power of Love inherent in the \textit{bhakti} path. Using them consistently, the negativity of the mind dissolves, allowing faith, Love and devotion to awaken in your heart.

As you evolve on the path, ever-increasing experiences of true unconditional Love and lasting happiness become more commonplace, regardless of the circumstances. Most often, this leads to a heartfelt yearning to connect more deeply with God, and Paramahamsa Vishwananda’s unconditional Love, support and guidance is available to everyone who wants to grow and evolve on this path to God-realization.


Additional practical support is also available through online programs, courses, retreats, and a global network of like-minded communities, known as \textit{saṅghas}.

Very naturally, as the fruits of this path blossom, you may find yourself caring more about the world around you. Such a shift leads to a growing desire to serve humanity and share the path of Divine Love with others, because, in the end, it is all about Love.

\noindent Just Love.

\noindent The greatest adventure of all.


\bigskip
\bigskip
\noindent {\avenirnextboldfont\large{Take the Next Step}}

\noindent Discover more at \textbf{bhaktimarga.org}